Curve counting in genus one: elliptic singularities {\it \&} relative geometry

### 2019-06-28 EllipticSingularitiesSplitting.tex
reintegrated some lemmas

We begin with the following lemma:
\begin{lemma} \label{lemma type C0 combinatorial types}
Let $\Dcal$ be a locus of type IIIb with combinatorial type $\Delta$ which contributes nontrivially to the Gromov--Witten invariant. For every nonzero vertex $\diamond$ of the tropical target, the fibre of the tropical map over $\diamond$ contains exactly one stable vertex, which lies on the radius.
\end{lemma}

@@

For each vertex of the target there must always be at least one vertex in the preimage which lies on the radius. This is implied by the fact that the combinatorial type $\Delta$ is a ray. In fact, we make the stronger assertion that the stable vertex mapping to $\diamond$ lies on the radius. The alternative is that a vertex dual to a trivial bubble lies on the radius. This latter condition implies that $\Dcal \to \Ecal$ has positive-dimensional fibres, so the contribution vanishes. 
\end{proof}

@@

Equipped with the strata logarithmic structures, the morphism
\[
\widetilde \Dcal\to \Ecal
\]
is a logarithmic morphism. The generic fiber of this morphism is toric variety.  \medskip

@@

This procedure induces a logarithmic modification $\widetilde\Ecal \to \Ecal$ which admits a modular description as a fibre product of genus zero centrally aligned maps coming from the vertices $\sqC_1,\ldots,\sqC_r$, and genus one aligned rubber maps satisfying fibre-direction factorisation from the vertex $\sqC_0$).

@@

\begin{lemma}
 Generically on $\Ecal$, $\widetilde \Dcal\to\widetilde\Ecal$ is given by an explicit projective bundle.
\end{lemma}

\begin{proof}
Let us assume that the curve does not degenerate, so that the dual graph is as prescribed by $\Delta$, in particular there are components $\sqC_1,\ldots\sqC_r$ generically lying on the circle of radius $\delta$; the moduli space of attachments for the elliptic singularity is the complement of the coordinate hyperplanes in $\PP(\bigoplus_{i=1}^rT_{q_i}C_i)$. On the boundary, the circle is allowed to miss some of the $\sqC_i$ - but not all of them! -, cutting the edge between them and the core (correspondingly, there is a destabilisation of the curve, and an elliptic singularity containing strictly semistable branches). Tropically, pushing some of the vertices off the circle creates new finite-length edges with coordinates $\ell_i\in\RR_{\geq_0}$ (see Figure \ref{fig:off_we_go}). Since not all of them can be strictly positive, this represents the fan of a projective space.
\end{proof}

@@

We said that $C_i^\prime$ is \emph{a} component of $C_i$ which lies on the radius. There may be more than one, but if this is the case then their tangent spaces are naturally identified. We obtain a natural morphism
\begin{equation*} \sum \operatorname{d}\!f \colon F(-\TT_{C_0}) \to \ev_q^\st \TT_H \end{equation*}
given by summing the derivatives at \emph{all} the inward-pointing nodes on the radius. {\bf \color{red} What does this next part mean?} Now, over $\widetilde\Dcal$ there is a natural section of the projective bundle $\PP(F) \to \widetilde\Dcal$ (obtained by projectivising the diagonal embedding). If we therefore lift the above morphism to the projectivisation, we obtain a composition: 
\begin{equation*} \OO_{\PP(F)}(-1) \to \pi^\st F \to \pi^\st(\ev_q^\st \TT_H (\TT_{C_0})).\end{equation*}
The vanishing locus of this section is given by an Euler class calculation on $\PP(F)$. In order to obtain $\Dcal$ we must then intersect this vanishing locus with $\OO_{\PP(F)}(1)$. %(which represents the natural section $\widetilde\Dcal \to \PP(F)$.

@@

{\bf \color{red} More needs to be said here.}

@@

>>>>

We begin with the following lemma:
\begin{lemma} \label{lemma type C0 combinatorial types}
Let $\Dcal$ be a locus of type IIIb with combinatorial type $\Delta$ which contributes nontrivially to the Gromov--Witten invariant. For every nonzero vertex $\diamond$ of the tropical target, the fibre of the tropical map over $\diamond$ contains exactly one stable vertex, which either is the core, or lies on the circle. Besides, the maximal contact order at the core is $m>1$.
\end{lemma}

@@

For each vertex of the target there must always be at least one vertex in the preimage which lies on the circle. This is implied by the fact that the combinatorial type $\Delta$ is a ray. If the vertex on the circle is strictly semistable - in particular the collapsed map $f_B$ is constant along the corresponing component -, it can be seen that the factorisation condition is unaffected by $\Gm$-scaling the coordinate corresponding to that vertex in the moduli space of attachments, so the map $\Dcal\to\Ecal$ has generic fiber of positive dimension.

In particular, notice that vertices lying on the circle and tropically mapped to $0\in\mathbb R_{\geq0}$ all have the same contact order at the core, because by the previous paragraph we may assume that only strictly semistable vertices are encountered on these paths. For the same reason, this contact order $m$ is maximal among those of the edges adjacent to the core.

Finally, in case the maximal contact order is $m=1$, the image tangent vectors must be linearly dependent, in which case we can deduce as above that factorisation will be satisfied indepedently of the choice of alignment.
\end{proof}

@@

Equipped with the strata logarithmic structures, the morphism
\[
\widetilde \Dcal\to \Ecal
\]
is a logarithmic morphism. The generic fiber of this morphism is a toric variety.  \medskip

@@

This procedure induces a logarithmic modification $\widetilde\Ecal \to \Ecal$ which admits a modular description as a fibre product of genus zero centrally aligned maps coming from the vertices $\sqC_1,\ldots,\sqC_r$, and genus one aligned rubber maps satisfying fibre-direction factorisation from the vertex $\sqC_0$.

@@

\begin{lemma}\label{lem:generic_proj_bundle}
 Generically on $\Ecal$, $\widetilde \Dcal\to\widetilde\Ecal$ is given by an explicit projective bundle.
\end{lemma}

\begin{proof}
Let us assume that the curve does not degenerate, so that the dual graph is as prescribed by $\Delta$, in particular there are components $\sqC_1,\ldots,\sqC_r$ generically lying on the circle of radius $\delta$; the moduli space of attachments for the elliptic singularity is the complement of the coordinate hyperplanes in $\PP(\bigoplus_{i=1}^rT_{q_i}C_i)$. On the boundary, the circle is allowed to miss some of the $\sqC_i$ - but not all of them! -, cutting the edge between them and the core (correspondingly, there is a destabilisation of the curve, and an elliptic singularity containing strictly semistable branches). Tropically, pushing some of the vertices off the circle creates new finite-length edges with coordinates $\ell_i\in\RR_{\geq_0}$ (see Figure \ref{fig:off_we_go}). Since not all of them can be strictly positive, this represents the fan of a projective space.

A slight modification of this argument is required to account for the fact that all vertices that lie on the circle and are mapped to $0\in\mathbb R_{\geq0}$ by the tropical map are generically aligned on $\Ecal$: indeed, by Lemma \ref{lemma type C0 combinatorial types}, they all have the same contact order $m$, therefore their $T_{q_i}C_i^{\otimes m}$ can be identified via the target tangent line on the fiber product \eqref{IIIa fibre product}, and, after saturation, the $T_{q_i}C_i$ themselves are canonically isomorphic on $\Ecal$. Thus, only one factor should be taken to represent all of them in the projective bundle above.
\end{proof}

@@

We said that $C_i^\prime$ is \emph{a} component of $C_i$ which lies on the radius. There may be more than one, but if this is the case then their tangent spaces are naturally identified.

The following construction allows us to give a canonical description of $\Dcal$ inside $\widetilde\Dcal$. Over $\widetilde\Dcal$ there is a natural section of the projective bundle $\PP(F) \to \widetilde\Dcal$ (obtained by projectivising the diagonal embedding). We may then pullback to $\widetilde\Dcal$ the map of vector bundles
\begin{equation}\label{eq:factorisation} \OO_{\PP(F)}(-1)(-\TT_{C_0}) \xrightarrow{\text{can}(-\TT_{C_0})} \pi^\st F(-\TT_{C_0}) \xrightarrow{\pi^*\sum\operatorname{d}f} \pi^\st(\ev_q^\st \TT_H),\end{equation}\footnote{We may assume the latter factors through $\ev_q^\st \TT_H\subseteq\ev_q^\st \TT_{\PP^N}$ by virtue of Lemma \ref{lemma type C0 combinatorial types}.}
whose vanishing tautologically expresses the condition that the stable map $f$ descend to the elliptic singularity associated to a point of $\widetilde\Dcal$. Thus, the class of $\Dcal$ can be represented via an Euler class computation. Note that the vector bundle $F$ can be explicitly realised by twisting the pullback of $\bigoplus_{i=1}^rT_{q_i}C_i$ from $\Ecal$ with a number of exceptional boundary corrections.

\begin{comment}
We obtain a natural morphism
\begin{equation*} \sum \operatorname{d}\!f \colon F(-\TT_{C_0}) \to \ev_q^\st \TT_H \end{equation*}
given by summing the derivatives at \emph{all} the inward-pointing nodes on the radius. {\bf \color{red} What does this next part mean?} Now, over $\widetilde\Dcal$ there is a natural section of the projective bundle $\PP(F) \to \widetilde\Dcal$ (obtained by projectivising the diagonal embedding). If we therefore lift the above morphism to the projectivisation, we obtain a composition: 
\begin{equation*} \OO_{\PP(F)}(-1) \to \pi^\st F \to \pi^\st(\ev_q^\st \TT_H (\TT_{C_0})).\end{equation*}
The vanishing locus of this section is given by an Euler class calculation on $\PP(F)$. In order to obtain $\Dcal$ we must then intersect this vanishing locus with $\OO_{\PP(F)}(1)$. %(which represents the natural section $\widetilde\Dcal \to \PP(F)$.

@@

{\bf \color{red} More needs to be said here.}
\end{comment}

@@

To establish a more direct connection with the geometry of the fibre product \eqref{IIIa fibre product}, recall that $\widetilde\Ecal$ has been built as an explicit modification of a finite cover of \eqref{IIIa fibre product}. For every $v\in\{\sqC_1,\ldots,\sqC_r\}$ - with the convention established at the end of Lemma \ref{lem:generic_proj_bundle} -, on $\widetilde\Ecal$ it makes sense to consider the line bundle $\OO(\delta_v)$ associated to a shortest path from the core to the circle in the $v$-direction. Construct $\mathcal P=\PP_{\widetilde\Ecal}(\bigoplus_{v\in \{\sqC_1,\ldots,\sqC_r\}}\OO(\delta_v))$; since $\sqC_1,\ldots,\sqC_r$ are always ordered on $\widetilde \Dcal$ (they may lie on the circle, be the first component off the circle, or degenerate and lie in the strict interior), there is a morphism $\widetilde \Dcal\to\mathcal P$ lifting $\widetilde \Dcal\to\widetilde\Ecal$. Consider the map of vector bundles on $\mathcal P$ (analogous to \eqref{eq:factorisation}):
\[ \OO_{\mathcal P}(-1)\to p^*(\bigoplus_{v\in \{\sqC_1,\ldots,\sqC_r\}}\OO(\delta_v))\xrightarrow{\sum\operatorname{d}f}\ev_{C_0}^*T_H.\]

The composite, regarded as a section $s$ of $G=\OO_{\mathcal P}(1)\otimes \operatorname{ev}_{C_0}^*(T_H)$, vanishes precisely where $f$ factors through the Smyth's singularity determined by the point in the fiber of $p$. Generically on $\widetilde\Ecal$, this is exactly what we are after - compare with Lemma \ref{lem:generic_proj_bundle}. On the boundary, though, it may happen that a simultaneous degeneration in the base and the fiber makes this factorisation trivial, in the sense that $\bar f$ would be constant on all the branches of the elliptic singularity.

\begin{lemma}
 For every $\Lambda_B\subseteq \{\sqC_1,\ldots,\sqC_r\}$ and $\Lambda_F=\{\sqC_1,\ldots,\sqC_r\}\setminus \Lambda_B$, $s$ vanishes along the loci where the factors corresponding to $\Lambda_B$ in $\widetilde \Ecal$ degenerate so that $f_i\colon C_i\to\PP^N$ is constant in a neighbourhood of $q_i$, and the fiber coordinates $(\lambda_j)_{j\in\Lambda_F}$ along $p\colon\mathcal P\to\widetilde \Ecal$ are set to be $0$.
\end{lemma}
Call these loci $\Xi_i,i\in I$ (compare with \cite[\S 3.2]{VZ} for an analogous phenomenon in the non-relative case); clearly, they may be in excess with respect to the expected codimension of $V(s)$, that is $N-1$. These loci are logarithmic substrata; therefore, there exists a logarithmic blow-up that principalizes them. We claim that $\widetilde\Dcal$ is (possibly a refinement of) such a blow-up. 
\begin{lemma}
 The pullback of $\Xi_i$ is principal in $\widetilde\Dcal$.
\end{lemma}
\begin{proof}
 Let us analyse the ideal of $\Xi_i$: it is generated by
 \begin{enumerate}[leftmargin=.5 cm]
  \item equations on $\widetilde\Ecal$ cutting the locus where $C_i$ degenerates so that $f_i$ is constant on a neighbourhood of $q_v$, $i\in \Lambda_B$: these correspond to the tropical functions $\delta_i-\operatorname{dist}(q_i,\circ)$;
  \item equations on the fiber of $p$ for the linear subspace corresponding to $j\in \Lambda_F$: tropically, these correspond to the $\ell_j$ discussed above.
 \end{enumerate}
By construction, on $\widetilde \Dcal$ it is always possible to tell which of these tropical lengths is the shortest (the circle we have drawn on $\Xi$ is degenerate - it passes through no vertex of positive horizontal degree -, therefore we need to decide for a strictly larger circle in order to lift to $\widetilde D$), which corresponds to principalising the ideal (see Figure \ref{fig:principalisation}).
\end{proof}

\begin{figure}
\includegraphics[scale=.5]{principalisation}
\caption{On the left, a generic point of $\mathcal P$. On the right, a point of $\Xi$. The dotted circle is the one we draw on $\mathcal P$. The continuous circle represents one of the possible choices of $\delta$ enlarging the dotted one (a stratum of $\widetilde\Dcal$).}\label{fig:principalisation}
\end{figure}

Finally, on $\widetilde\Dcal$ (or, in fact, on an intermediate blow-up $\widetilde{\mathcal P}\to\mathcal P$) there are exceptional divisors $\widetilde \Xi_i$ corresponding to $\Xi_i$ in $\mathcal P$. Since the $\Xi_i$ exhaust all the excess-dimensional components of the vanishing locus of $s\in\Gamma(\mathcal P,G)$, the latter induces a section $\tilde s\in\Gamma(\widetilde\Dcal,G(-\sum_i\widetilde\Xi_i))$ by pullback and twisting, whose vanishing locus is dimensionally transverse to the boundary, and whose open part coincides with the generic point of $\Dcal$, therefore $V(\tilde s)=\Dcal$.

Expressing the pushforward of the Chern classes of $\OO_F(1)$ and $\OO(\widetilde\Xi_i)$ in terms of tautological classes on the fiber product underlying $\Ecal$, we get:
\begin{cor}
 After pushforward to $\Ecal$, the contribution of $\Dcal$ to the invariants can be computed in terms of tautological integrals on moduli spaces of maps with lower numerics.
\end{cor}
